---
title: 'Reproducible Research: Peer Assessment 1'
author: "Anurati Chatty"
date: "10/01/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## R Markdown for peer review 
This is an R markdown generated for the Reproducible Research peer review assesment 1. 

## Loading and preprossesing the data
Load the data and store it in activity
```{r}
library(dplyr)
activity <- read.csv("activity.csv")
head(activity)
activity$date <- as.Date(activity$date,"%Y-%m-%d")
```

## What is mean total number of steps taken per day?
First we calculate the totaal number of steps taken per day using tapply

```{r}
steps_day <- tapply(activity$steps,activity$date,sum,na.rm=TRUE)
```
Second plotting the histogram for number of steps taken per day

```{r}
hist(steps_day, xlab = "number of steps", ylab="frequency", main="Histogram of number of steps taken in a day")
```
Barplot

```{r}
barplot(height = steps_day, xlab = "date", ylab = "number of steps", main = "total number of steps taken in a day")
```
Mean of total number of steps taken per day

```{r}
mean(steps_day)
```
Median of total number of steps taken per day

```{r}
median(steps_day)
```
## What is the average daily activity pattern?
Making a time series plot of 5 minute intervals and the average number of steps taken, averaged across all days

```{r}
day_mean <- tapply(activity$steps,activity$interval , mean, na.rm=TRUE)
head(day_mean)
plot(names(day_mean), day_mean, type = "l", xlab = "5 min intervals", ylab = "no of steps", main = "average number of steps taken in a time interval")
```
 5 minute interval containing the maximum average number of steps across all days
 
```{r}
day_mean[which(day_mean==max(day_mean))]
```
## Imputing missing values
Calculating the total number of missing values

```{r}
totalna <- sapply(activity$steps,is.na)
sum(totalna)
```
Filling all the missing values with the average number of steps taken for that 5 min interval and creating a new dataset which is identical to the original dataset with missing values filled

```{r}
activity_new <- activity
for( i in 1:length(activity_new$steps)){
  if(is.na(activity_new$steps[i])){
    activity_new$steps <- replace(activity_new$steps,i,day_mean[as.character(activity_new$interval[i])])
  }
}
head(activity_new)
```
Histogram for the total number of steps taken everyday for the new dataset
```{r}
steps_day_new <- tapply(activity_new$steps,activity_new$date,sum,na.rm=TRUE)
hist(steps_day_new, xlab = "number of steps", ylab="frequency", main="Histogram of number of steps taken in a day")
```
Barplot

```{r}
barplot(height = steps_day_new, xlab = "date", ylab = "number of steps", main = "total number of steps taken in a day")
```
Mean of steps taken everyday for the new dataset

```{r}
mean(steps_day_new)
```
Median of steps taken everyday for the new dataset

```{r}
median(steps_day_new)
```
Therefore we can conclude that the mean and median are higher when the missing values are filled with the average number of steps for that 5 minute interval

##Are there differences in activity patterns between weekdays and weekends?
We use the weekdays function to create a new factor called day with values "Weekday" and "Weekend"

```{r}
activity_new$day <- mutate(activity_new, day=weekdays(activity_new$date))
weekdays <- c("Monday","Tuesday","Wednesday","Thursday","Friday")
activity_new$day <- factor((weekdays(activity_new$date) %in% weekdays), 
                       levels=c(FALSE, TRUE), labels=c('Weekend', 'Weekday'))
```
Making a panel plot containing a time series plot of the 5 minute interval and the average number of steps taken in a day across all the weekdays and weekends.

```{r}
weekday <- subset(activity_new,day=="Weekday")
weekend <- subset(activity_new, day=="Weekend")

steps_weekday_avg <-tapply(weekday$steps, weekday$interval, mean)
steps_weekend_avg <-tapply(weekend$steps, weekend$interval, mean)

par(mfrow=c(2,1), mar=c(4,4,2,1))
plot(names(steps_weekday_avg), steps_weekday_avg, type = "l", xlab = "5 min intervals in weekdays", ylab = "no of steps", main = "average number of steps taken in a time interval on a weekday")
plot(names(steps_weekend_avg), steps_weekend_avg, type = "l", xlab = "5 min intervals in weekends", ylab = "no of steps", main = "average number of steps taken in a time interval on a weekend")

```